Clustering SCPCL000822 and SCPCL000824

### 2024-08-02 analyses/cell-type-ewings/exploratory_analysis/05-cluster-exploration.Rmd
update plots based on review

# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap()
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

@@

```{r, assign_clusters_functions}
# get clusters for a specified value of k
get_clusters <- function(pcs, k) {
  clusters <- bluster::clusterRows(
    pcs,
    bluster::NNGraphParam(
      k = k,
      type = "jaccard",
      cluster.fun = "louvain"
    )
  )

@@

```{r, plotting_functions}
# heatmap comparing cluster assignments to SingleR labels
cluster_celltype_heatmap <- function(cluster_classification_df,
                                     id) {
  # get a jaccard mtx for each cluster param
  jaccard_df_list <- cluster_classification_df |>
    dplyr::filter(library_id == id) |>
    split(cluster_classification_df$nn_param) |>
    purrr::map(\(df) {
      make_jaccard_matrix(
        df,
        "cluster",
        "singler_lumped"
      )
    })

  # turn into heatmap list
  make_heatmap_list(jaccard_df_list, column_title = glue::glue("{id}-clusters"), legend_match = "k_5")
}

@@

```{r}
# perform clustering across k = 5-40 with increments of 5
all_cluster_results <- sce_list |>
  purrr::map(cluster_sweep) |>
  dplyr::bind_rows(.id = "library_id")
```

@@

```{r, fig.height=15, fig.width=10}
# look at clustering results across all params
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.1) +
  facet_grid(
    cols = vars(library_id),
    rows = vars(nn_param)
  ) +
  theme(aspect.ratio = 1)
```

@@

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats_df <- sce_list |>
  purrr::imap(\(sce, library_id){
    get_cluster_stats(sce, library_id, all_cluster_results)
  }) |>
  dplyr::bind_rows(.id = "library_id")
```

```{r}
# summarize the silhouette width and purity for each nn_param used
summary_stats_df <- all_stats_df |>
  dplyr::group_by(library_id, nn_param) |>
  dplyr::summarise(
    mean_width = mean(width),
    mean_purity = mean(purity)
  ) |>
  dplyr::mutate(
    # make sure the order is correct
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r}
# silhouette width for different params
ggplot(summary_stats_df, aes(x = nn_param, y = mean_width)) +
  geom_point() +
  facet_wrap(vars(library_id),
    scales = "free_y"
  ) +
  labs(y = "Average silhouette width")
```

```{r}
# cluster purity for different params
ggplot(summary_stats_df, aes(x = nn_param, y = mean_purity)) +
  geom_point() +
  facet_wrap(vars(library_id),
    scales = "free_y"
  ) +
  labs(y = "Average cluster purity")
```

@@

# plot stability across all values of k
ggplot(stability_df, aes(x = k_value, y = ari)) +
  geom_jitter(width = 0.1) +
  facet_grid(rows = vars(library_id)) +
  labs(title = "Cluster stability")
```

@@

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(classification_df, by = c("library_id", "barcodes")) |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )
```

```{r, fig.height=10, fig.width=7}
# get heatmap for each library showing cluster assignments vs. singler assignments
names(sce_list) |>
  purrr::map(\(id){
    cluster_celltype_heatmap(
      cluster_classification_df,
      id
    )
  })
```

@@

```{r, fig.height=10, fig.width=5}
# plot to look at marker gene expression for all cell types across all clusters
# choose k=25 based on previous stats
names(sce_list) |>
  purrr::map(\(id) {
    plot_marker_genes(cluster_classification_df,
      k_value = "k_25",
      id
    )
  })
```

@@

- Any clusters with high tumor marker gene expression will be labeled "tumor". 
- Any clusters with high immune marker gene expression also correspond to clusters that are predominantly "macrophage" in the heatmaps, so they will be labeled as "macrophage". 
- Any clusters with high endothelial marker gene expression also correspond to clusters that are predominantly "endothelial cell" in the heatmaps, so they will be labeled as "macrophage". 
- There is one cluster in `SCPCL000822` that has high mesenchymal gene expression and is predominantly chondrocytes so we will label this cell as "chondrocytes". 
- All other clusters with high mesenchymal gene expression appear to have a mix of chondrocytes, fibroblasts, and smooth muscle cells so we will just label these as "mesenchymal-like cells".

@@

```{r}
# look at cell type assignments on the UMAP
ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = cluster_celltype)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_grid(cols = vars(library_id)) +
  theme(aspect.ratio = 1) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
  scale_color_brewer(palette = "Dark2")
```

@@

```{r}
# grab columns that contain marker gene sums
marker_gene_columns <- colnames(cluster_umap_df)[which(endsWith(colnames(cluster_umap_df), "_sum"))]

# create individual umaps and combine into one
marker_gene_columns |>
  purrr::map(\(column){
    genelist_name <- stringr::str_remove(column, "_sum")

    ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = !!sym(column))) +
      geom_point(alpha = 0.5, size = 0.5) +
      facet_grid(cols = vars(library_id)) +
      theme(aspect.ratio = 1) +
      scale_color_viridis_c() +
      labs(
        title = genelist_name
      )
  })
```

>>>>

# source in helper functions: plot_density() and calculate_sum_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

@@

```{r, assign_clusters_functions}
# get louvain, jaccard clusters for a specified value of k (nearest neighbors)
get_clusters <- function(pcs, k) {
  clusters <- bluster::clusterRows(
    pcs,
    bluster::NNGraphParam(
      k = k,
      type = "jaccard",
      cluster.fun = "louvain"
    )
  )

@@

```{r, plotting_functions}
# plot individual stats for clusters, either purity or width
plot_cluster_stats <- function(all_stats_df,
                               stat_column,
                               id) {
  plot_df <- all_stats_df |>
    dplyr::filter(library_id == id)

  ggplot(plot_df, aes(x = nn_param, y = {{ stat_column }})) +
    # ggforce::geom_sina(size = .2) +
    ggbeeswarm::geom_quasirandom(method = "smiley", size = 0.1) +
    stat_summary(
      aes(group = nn_param),
      color = "red",
      # median and quartiles for point range
      fun = "median",
      fun.min = function(x) {
        quantile(x, 0.25)
      },
      fun.max = function(x) {
        quantile(x, 0.75)
      }
    )
}

# heatmap comparing cluster assignments to SingleR labels
cluster_celltype_heatmap <- function(cluster_classification_df,
                                     id) {
  # get a jaccard mtx for each cluster param
  jaccard_df_list <- cluster_classification_df |>
    dplyr::filter(library_id == id) |>
    split(cluster_classification_df$nn_param) |>
    purrr::map(\(df) {
      make_jaccard_matrix(
        df,
        "cluster",
        "singler_lumped"
      )
    })

  # turn into heatmap list
  make_heatmap_list(jaccard_df_list, column_title = glue::glue("{id}-clusters"), legend_match = "k_5", cluster_rows = FALSE)
}

@@

```{r}
# perform clustering across k = 5-40 with increments of 5
all_cluster_results <- sce_list |>
  purrr::map(cluster_sweep) |>
  dplyr::bind_rows(.id = "library_id") |>
  dplyr::mutate(
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

@@

```{r, fig.height=15, fig.width=10}
# look at clustering results for each library across all params
library_ids |>
  purrr::map(\(id){
    umap_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = cluster)) +
      geom_point(alpha = 0.5, size = 0.1) +
      facet_wrap(vars(nn_param)) +
      theme(
        aspect.ratio = 1
      ) +
      labs(title = id) +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))
  })
```

@@

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats_df <- sce_list |>
  purrr::imap(\(sce, library_id){
    get_cluster_stats(sce, library_id, all_cluster_results)
  }) |>
  dplyr::bind_rows(.id = "library_id") |>
  dplyr::mutate(
    # make sure the order is correct
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r}
# silhouette width for different params
library_ids |>
  purrr::map(\(id){
    plot_cluster_stats(all_stats_df, width, id)
  })
```

```{r}
# cluster purity for different params
library_ids |>
  purrr::map(\(id){
    plot_cluster_stats(all_stats_df, purity, id)
  })
```

@@

# plot stability across all values of k
ggplot(stability_df, aes(x = k_value, y = ari)) +
  geom_jitter(width = 0.1) +
  facet_grid(rows = vars(library_id)) +
  labs(title = "Cluster stability") +
  stat_summary(
    aes(group = k_value),
    color = "red",
    # median and quartiles for point range
    fun = "median",
    fun.min = function(x) {
      quantile(x, 0.25)
    },
    fun.max = function(x) {
      quantile(x, 0.75)
    }
  )
```

@@

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(classification_df, by = c("library_id", "barcodes")) |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf),
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r, fig.height=10, fig.width=7}
# get heatmap for each library showing cluster assignments vs. singler assignments
library_ids |>
  purrr::map(\(id){
    cluster_celltype_heatmap(
      cluster_classification_df,
      id
    )
  })
```

@@

```{r, fig.height=10, fig.width=5}
# plot to look at marker gene expression for all cell types across all clusters
# choose k=25 based on previous stats
library_ids |>
  purrr::map(\(id) {
    plot_marker_genes(cluster_classification_df,
      k_value = "k_25",
      id
    )
  })
```

@@

- Any clusters with high tumor marker gene expression will be labeled "tumor". 
- Any clusters with high immune marker gene expression also correspond to clusters that are predominantly "macrophage" in the heatmaps, so they will be labeled as "macrophage". 
- Any clusters with high endothelial marker gene expression also correspond to clusters that are predominantly "endothelial cell" in the heatmaps, so they will be labeled as "macrophage". 
- There is one cluster in `SCPCL000822` that has high mesenchymal gene expression and is predominantly chondrocytes so we will label this cell as "chondrocytes". 
- All other clusters with high mesenchymal gene expression appear to have a mix of chondrocytes, fibroblasts, and smooth muscle cells so we will just label these as "mesenchymal-like cells". 
- Any clusters that do not have high expression of any marker gene lists are labeled as "unknown".
This corresponds to cluster 9 and 12 in `SCPCL000824`.

@@

```{r, fig.height=7, fig.width=7}
# look at cell type assignments on the UMAP
ggplot(cluster_umap_df, aes(x = UMAP1, y = UMAP2, color = cluster_celltype)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_grid(cols = vars(library_id)) +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom"
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5))) +
  scale_color_brewer(palette = "Dark2")
```

@@

```{r, fig.height=7, fig.width=7}
# grab columns that contain marker gene sums
marker_gene_columns <- colnames(cluster_umap_df)[which(endsWith(colnames(cluster_umap_df), "_sum"))]

# for each sample show a faceted umap by cell type
# color is gene expression of all marker genes for that cell type
library_ids |>
  purrr::map(\(id){
    plot_df <- cluster_umap_df |>
      dplyr::filter(library_id == id) |>
      tidyr::pivot_longer(
        cols = all_of(marker_gene_columns),
        names_to = "cell_type",
        values_to = "gene_expression"
      ) |>
      dplyr::mutate(cell_type = stringr::str_remove(cell_type, "_sum"))

    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
      geom_point(alpha = 0.5, size = 0.2) +
      facet_wrap(vars(cell_type)) +
      theme(aspect.ratio = 1) +
      scale_color_viridis_c() +
      labs(
        title = id
      )
  })
```
